\Large Fast and adaptive cointegration based model for forecasting high frequency financial time series

### 2016-04-18 articles-conferences/iscef2016/ISCEF2016-slides/iscef2016-slides.tex
Inicio articulo 2016
version final ISCEF 2016

\begin{frame}
\frametitle{Outline}
\begin{itemize}
\item Motivation: economic relations
\item Cointegration and Integration concepts
\item VECM model
\item Proposal
\item Experiments
\item Conclusions
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Motivation}
Financial time series are mainly non-stationary.  However, one or more linear
combinations of non-stationary time series can be stationary even though
individually they are not.

When this happens it is said they have a long-run
equilibrium relationship and the variables are \textbf{cointegrated}.
\end{frame}

\section{Background}
\subsection{Definitions}

\begin{frame}
\frametitle{Integration }
\begin{block}{Integration of order $d$}
A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
differentiating the variable $d$ times, we get an stationary process, more precisely:
\[
(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
\]
\noindent where I(0) is a stationary time series and $L$ is the lag operator:
\[
(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
\]
\end{block}
\end{frame}

\subsection{Integration example}
\begin{frame}
\frametitle{Financial Time Series}
\begin{columns}
\column[t]{5cm}
Financial time series are commonly a I(1) process 
\begin{center}
\includegraphics[width=0.85\textwidth]{img/EURUSD}
\end{center}
\column[t]{5cm}
Since their differences are stationary (I(0) process).
\begin{center}
\includegraphics[width=0.85\textwidth]{img/DEURUSD}
\end{center}
\end{columns}
\end{frame}

@@

\section{Proposal}
\subsection{Motivation}
\begin{frame}
\frametitle{VECM difficulties}
\begin{itemize}
\item VECM assumes cointegration vectors do not change in time. However, these might change due to several economic factors.
\item To get updated cointegration vectors in time is computationally expensive because 
the Johansen method and OLS are used to obtain them.
\end{itemize}
\end{frame}

@@

\begin{frame}
\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
\begin{itemize}
\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{AVECM Algorithm}
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$j$: Starting point of testing \\
$it$: Ending point of testing \\
$ps$: list of $p$ values \\
$Ls$: list of $L$ values ($L<N$) \\
$m$: Iterations to determine parameters ($m < N-L$)\\
\ENSURE  $\,$ \\
$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
\FOR { $i =j$ to $it$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
\ENDFOR
\end{algorithmic}
\end{frame}

@@

\begin{frame}
\frametitle{Evaluation methods}
\begin{block}{MSE}
Measures the distance between forecasts
and the true values and large deviations from the true value have a
large impact due to squaring forecast error.
{\color{blue}
\begin{equation}\label{eq:MSE}
\text{MSE} = 
\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
\end{equation}}
\end{block}
\begin{block}{\bf $U$-statistic}
Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
\end{block}
\end{frame}

\section{Experiments}
\subsection{Uno}
\begin{frame}
\frametitle{Data}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from Dukascopy, a free
database which gives access to the Swiss Foreign Exchange marketplace.
The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{frame}

\begin{frame}
\frametitle{Unit root tests}
VECM can only be used if the time series are I(1). Therefore, 
before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\end{frame}

\begin{frame}
\frametitle{Performance accuracy}
\small
\begin{tabular}{ccccccc}
& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 & AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
 \end{tabular}
 \end{frame}
%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.

\begin{frame}
\frametitle{Parallel implementation}
The function get\_best\_params($Ls,ps,m,\mathbf{Y}$) was implemented using MPI in Python. MPI allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
\end{frame}

\begin{frame}
\frametitle{Parameters setting}
\begin{itemize}
\item Parameter $it=100$
\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. 
\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
\end{itemize}
\end{frame}

@@

\section{Conclusions}
\subsection{Uno}
\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/} and in the last samples.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
\item Increasing the number of parameters will always lead to
better performance measures.
\end{itemize}
\end{frame}

>>>>

%\begin{frame}
%\frametitle{Outline}
%\begin{itemize}
%\item Motivation: economic relations
%\item Cointegration and Integration concepts
%\item VECM model
%\item Proposal
%\item Experiments
%\item Conclusions
%\end{itemize}
%\end{frame}

@@

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item Financial time series are mainly non-stationary.  However, one or more linear
combinations of non-stationary time series can be stationary even though
individually they are not. When this happens it is said they have a long-run
equilibrium relationship and the variables are \textbf{cointegrated}.
\item The Vector Error correction model (VECM) is used to model cointegrated time series but is used only with batch data and rarely used with high frequency data.
\end{itemize}
\end{frame}

\subsection{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{}
To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
\end{block}
\end{frame}

\section{Background}
%
%\begin{frame}
%\frametitle{Integration }
%\begin{block}{Integration of order $d$}
%A time series $\mathbf{y}$ is said to be integrated of order $d$ if after
%differentiating the variable $d$ times, we get an stationary process, more precisely:
%\[
%(1-L)^d \mathbf{y} \sim \text{I(0)} \, ,
%\]
%\noindent where I(0) is a stationary time series and $L$ is the lag operator:
%\[
%(1-L)\mathbf{y} = \Delta \mathbf{y}=\mathbf{y}_t  -\mathbf{y}_{t-1} \quad \forall t
%\]
%\end{block}
%\end{frame}
%
%\subsection{Integration example}
%\begin{frame}
%\frametitle{Financial Time Series}
%\begin{columns}
%\column[t]{5cm}
%High frequency financial time series are commonly an I(1) process 
%\begin{center}
%\includegraphics[width=0.85\textwidth]{img/EURUSD}
%\end{center}
%\column[t]{5cm}
%Since their differences are stationary (I(0) process).
%\begin{center}
%\includegraphics[width=0.85\textwidth]{img/DEURUSD}
%\end{center}
%\end{columns}
%\end{frame}

@@

\section{Proposal}
\subsection{Motivation}
\begin{frame}
\frametitle{VECM difficulties in an online context}
\begin{itemize}
\item VECM assumes cointegration vectors do not change in time. However, these might change due to several economic factors.
\item To get updated cointegration vectors in time is computationally expensive because 
the Johansen method and OLS are used to obtain them. It would be good to know how the number of cointegration vectors are affected by the number of data used $L$ and the number of lags of VECM $p$.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Data}
\begin{itemize}
\item Tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
\item This data was collected from Dukascopy, a free
database which gives access to the Swiss Foreign Exchange marketplace.
\item The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. 
%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{itemize}
\end{frame}

@@

%\begin{frame}
%\frametitle{AVECM Algorithm}
%\begin{algorithmic}[1]
%\REQUIRE $\,$ \\
%$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
%$j$: Starting point of testing \\
%$it$: Ending point of testing \\
%$ps$: list of $p$ values \\
%$Ls$: list of $L$ values ($L<N$) \\
%$m$: Iterations to determine parameters ($m < N-L$)\\
%\ENSURE  $\,$ \\
%$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
%\FOR { $i =j$ to $it$ }
%   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
%    \STATE $L,p \gets
%    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
%        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
%        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
%\ENDFOR
%\end{algorithmic}
%\end{frame}

@@

\begin{frame}
\frametitle{Results}
\begin{tabular}{ccccccc}

& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 &
AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.07 e-09 & 1.14 e-09 &  9.25 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.66 e-09 & 1.74 e-09 &  6.95 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.85 e-10 & 6.35 e-10 &  2.89 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.34 e-06 & 6.51 e-06 &  6.85 e-05 & & 0.6964 & 0.7057\\
 \end{tabular}
\end{frame}

%\begin{frame}
%\frametitle{Evaluation methods}
%\begin{block}{MSE}
%Measures the distance between forecasts
%and the true values and large deviations from the true value have a
%large impact due to squaring forecast error.
%{\color{blue}
%\begin{equation}\label{eq:MSE}
%\text{MSE} = 
%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
%\end{equation}}
%\end{block}
%\begin{block}{\bf $U$-statistic}
%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
%\end{block}
%\end{frame}

\section{Experiments}
\subsection{Uno}

%\begin{frame}
%\frametitle{Unit root tests}
%VECM can only be used if the time series are I(1). Therefore, 
%before running the tests, we firstly checked whether the time series were
%I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
%\end{frame}
%
%\begin{frame}
%\frametitle{Performance accuracy}
%\small
%\begin{tabular}{ccccccc}
%& \multicolumn{3}{l}{MSE} & &
%\multicolumn{2}{c}{$U$-statistic} \\ 
%\hhline{~---~--} \\
% & AVECM & ARIMA & p-value & &
%AVECM & ARIMA \\ 
%
%\hline
% EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
% GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
% USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
% USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
% \end{tabular}
% \end{frame}
%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.

\begin{frame}
\frametitle{Parallel implementation}
The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
\end{frame}

\begin{frame}
\frametitle{Parameters setting}
\begin{itemize}
%\item Parameter $it=100$
\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. 
\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
\end{itemize}
\end{frame}

@@

\section{Conclusions}
\subsection{Uno}
\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
\item Increasing the number of parameters will always lead to
better performance measures.
\end{itemize}
\end{frame}
